Collaborative Visual Analysis on the Web with RCloud

### 2015-06-25 conclusion.tex
fixups

\section{Conclusions and Future Work}
We designed. implemented and deployed a prototype visual analytics environment
for data exploration, sharing, presentation, and publishing. It is a step toward
practical ``DevOps for data science'' and reproducible, publishable data
science experiments.

>>>>

\section{Conclusions and Future Work}
We designed, implemented and deployed a prototype visual analytics environment
for data exploration, sharing, presentation, and publishing. It is a step toward
practical ``DevOps for data science'' and reproducible, publishable data
science experiments.

### 2015-06-25 discussion.tex
fixups

\paragraph*{Curation} Even without formal group management, users may
curate groups of related notebooks using notebook tree folders.
We found this particularly effective in collecting and distributing
training materials.

@@

\paragraph*{Versioning and Upgrading}
Some important aspects of the environment are more difficult to manage
in RCloud than in conventional systems. RCloud does not separate the
development and deployment environments, and every version of every
prototype is shared by default. Although this encourages
collaboration, it also quickly exposes errors such as
misconfiguration, mismatches between versions of packages, and
programming errors that can unintentionally affect production
websites. For example, if the RCloud website is distributed across
servers in different machines, different versions of system software
can cause errors that are hard to find and fix.

Versioning of different software pieces must be managed in several
levels: in data analysis and visualization scripts (that is, the
notebooks themselves); in the R environment such as the installed
libraries and packages; and in the external environment such as the
operating system, libraries, and protocols spoken by remote
services. At one end of the scale, we have full control over the
versions via git, along with conventions that fix stable notebook
versions.  The R environment itself is under the control of its
package manager that has rules to ensure reasonable consistency.  It
is at least possible for RCloud to access information about package
versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not
strong. At the other end of the scale, there is not much reason to
expect version control for the external environment, and consider what
would happened if a web service that is used by a notebook changes the
format of its output: notebooks now stop working, but nothing in
RCloud has changed.

@@

\paragraph*{Caching}
An important consideration is how and where to introduce caching in
RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is otherwise transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Currently, RCloud offers no support for implicit caching, but since it
supports the entirety of the R language, it's possible to incorporate
caching in R libraries. Still, since the reuse of intermediate
results is a recurring problem in data exploration environments, it
seems better to implement this capability in a general purpose
associative cache. We envision cache management being implemented in a
new middle layer to be added in the future.

\paragraph*{Security}
Because RCloud allows arbitrary R code to be run
\emph{by design}, it's not straightforward to protect deployments
from running unauthorized arbitrary code. RCloud uses an object capability model
\cite{Miller:2006:RCT} recently added to the Rserve
protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
to remote procedure calls; this way, web browsers never directly instruct the
RCloud backend to execute arbitrary code, preventing unauthenticated clients
from making unauthorized calls to the RCloud runtime environment.
%
Our back-end environment, on the other hand, assumes a high degree of trust between users:
access control for information security is delegated to the host
operating system and web server. In practice, operating
systems rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome. This means that information security
in RCloud is a delicate, unsolved matter. More
sophisticated approaches, such as information flow analysis,
and query languages that ensure differential privacy, are
still research topics~\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

\paragraph*{Nonforgetful, effortless exploration is hard}
Users reported that RCloud's combined notebook and command line
interface interferes with their familiar use of R for exploration. In
current R practice, the command prompt is used to try things out. It
is very fast and there is often no commitment: nothing gets saved by
default, but clean up is automatic. In contrast, in RCloud's
notebook interface, cells generated doing exploration are saved in the
notebook. We arrived here from an insistence on reproducibility: it is a
core principle of RCloud that no command should be run without being
saved. In our experience some history steps are only found to be
important in hindsight, and deleting history by default makes it too
easy to lose those steps by accident.

Surprisingly, we found that when users realize that \emph{scratch work
is getting saved}, the extra versions become a mental burden for
them. Clearly, RCloud is an extreme, and the right solution is
in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
data artifacts to the processes that generated them. As a result,
bookkeeping can be reduced to selecting the \emph{data} to be kept,
and the code which generates it can be derived and saved transparently.

@@

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, a global notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever. In addition, important metadata (has this notebook been
recently run, or edited?) is also not easily searchable. In other words,
\emph{textual search} is easy, but \emph{relevance search} is what we
need. Allowing users to \emph{tag} notebooks is natural, but it
relies on users making a conscious effort, which goes against
our ideal of not asking the user to do work for the system.
It is better to have the system work for the user,
relying on passive metrics, such as how often a notebook has been run.
More generally, infrastructure for automatic \emph{curation} of
software artifacts generated by teams seems to be an important avenue
for future work.

>>>>

\paragraph*{Curation} Even without formal group management, users may
organize groups of related notebooks using tree folders.
We found this particularly effective in collecting and distributing
training materials.

@@

\paragraph*{Versioning and Upgrading}
Some aspects of the programming environment are more difficult to
manage in RCloud than in conventional systems. RCloud does not separate
the development and deployment environments, and every version of every
prototype is shared by default. While this encourages
collaboration, it also quickly exposes errors such as
misconfiguration, mismatches between package versions, and
programming errors, that can unintentionally affect production
websites. For example, if an RCloud website is distributed across
multiple servers, conflicting versions of system software
can cause errors that are hard to debug.

Versioning of software components must be managed in several
levels: in data analysis and visualization scripts (the
notebooks themselves); in the R environment such as the installed
libraries and packages; and in the external environment such as the
operating system, libraries, and protocols spoken by remote
services. At one end of the scale, we have full control over notebooks
via git, along with conventions to fix stable notebook versions.
The R environment is under the control of its package manager that
has rules to ensure reasonable consistency.
(It is possible for RCloud to access information about package
versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not
strong.) At the other end of the scale, there is hardly any version
control for the external environment, and if an external service 
used by a notebook changes its API, the notebook can stop working.

@@

\paragraph*{Caching}
An important issue is how and where to introduce caching in
RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Currently, RCloud offers no support for implicit caching, but 
it's possible to incorporate caching in individual R libraries.
The reuse of intermediate results is a recurring problem
in many data exploration environments.
We envision general cache management being implemented in a
new middle layer in the future.

\paragraph*{Security}
Because RCloud allows arbitrary R code to be run
\emph{by design}, it's not straightforward to protect deployments
from running unauthorized arbitrary code. RCloud uses an object capability model
\cite{Miller:2006:RCT} recently added to the Rserve
protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
to remote procedure calls; web browsers never directly instruct the
RCloud backend to execute arbitrary code, preventing unauthenticated clients
from making unauthorized calls to the RCloud runtime environment.
%
Our back-end environment, on the other hand, assumes a high degree of trust between users:
access control for information security is delegated to the host
operating system and web server.
Most operating systems rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome. This means that information security
in RCloud is a delicate, unsolved matter. More
sophisticated approaches, such as information flow analysis,
and query languages that ensure differential privacy, are
open research topics~\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

\paragraph*{Nonforgetful, effortless exploration is hard}
Users reported that RCloud's combined notebook and command line
interface interferes with their familiar use of R for exploration. In
current R practice, the command prompt is used to try things out. It
is very fast and there is often no commitment: nothing gets saved by
default, and clean up is automatic. In contrast, in RCloud's
notebook interface, cells generated doing exploration are saved in the
notebook. We arrived here from an insistence on reproducibility: it is a
core principle that no command should be run without being
saved. In our experience some history steps are only found to be
important in hindsight, and deleting history by default makes it too
easy to lose those steps by accident.

Surprisingly, we found that when users realize that \emph{scratch work
is getting saved}, the extra versions become burdensome to them.
Clearly, RCloud is an extreme, and the right solution is
in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
data artifacts to the processes that generated them. 
Bookkeeping would be reduced to selecting the \emph{data} to be kept,
and the code that generated it would be derived and saved transparently.

@@

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, a global notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever. In addition, important metadata (has this notebook been
recently run, or edited?) is not easily searchable. In other words,
\emph{textual search} is easy, but \emph{relevance search} is what we
need. Allowing users to \emph{tag} notebooks is natural, but
relies on users making a conscious effort, which goes against
our ideal of not asking the user to do work for the system.
% It is better to have the system work for the user,
% relying on passive metrics, such as how often a notebook has been run.
% More generally, infrastructure for automatic
Automatic
\emph{curation} of
software artifacts generated by teams seems to be an important avenue
for future work.

### 2015-06-25 paper.tex
submitted
figure squeezing
fixups

%\documentclass[journal]{vgtc}                % final (journal style)
\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

@@

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{Stephen North and Carlos Scheidegger and Simon Urbanek and
  Gordon Woodhull}

\authorfooter{
%% insert punctuation at end of each item
}

@@

%% Abstract section.
\abstract{ Consider the emerging role of data science teams embedded in
  larger organizations. Individual analysts work on loosely related problems,
  and must share their findings with each other and the organization at large,
  moving results from exploratory data analyses (EDA)
  into automated visualizations, diagnostics and reports deployed for wider consumption.
  There are two problems with the current
  practice. First, there are gaps in this workflow: EDA is performed
  with one set of tools, and automated reports and deployments with another.
  Second, these environments often assume a single-developer perspective,
  while data scientist teams could get much benefit from easier sharing
  of scripts and data feeds, experiments, annotations, and automated
  recommendations, which are well beyond what traditional version control
  systems provide. We contribute and justify the following three
  requirements for systems built to support current data science teams
  and users: \emph{technology transfer}, \emph{coexistence}, and
  \emph{discoverability}. In addition, we contribute the design and
  implementation of RCloud, a system that supports the requirements of
  collaborative data analysis, visualization and web deployment.
  The biggest deployment of RCloud has been in active use for more than two years,
  and has about fifty active users. We report on interviews with some of these users, and
  discuss the design decisions, tradeoffs and limitations, comparing RCloud
  to other current proposals.
} % end of abstract

@@

%% Uncomment below to include a teaser figure.
\teaser{
  \centering
  \includegraphics[width=.95\linewidth]{fig/teaser/teaser.pdf}
  % \vspace{-.5em}
  \caption{
    An overview of features in RCloud. RCloud supports an environment in which a large number of loosely-related problems in data and visual analytics are solved by a small number of \emph{hackers} and \emph{scripters}. In such a shared environment, \emph{discoverability} is a concern. RCloud supports search, annotation, recommendation, and commenting for all notebooks, and provides an overview where users can \emph{browse} popular and recent analyses. When problems and data sources change frequently, deployment can become very costly; RCloud supports \emph{transparent and automatic} deployment of analyses as web pages and web services, allowing a seamless transition from data exploration work to production web services.
  }
% \vspace{-.5em}
}

>>>>

\documentclass[journal]{vgtc}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

@@

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{341}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{Stephen North, Carlos Scheidegger, Simon Urbanek and
  Gordon Woodhull}

\authorfooter{
%% insert punctuation at end of each item
\item 
Stephen North is with Infovisible LLC, Oldwick, USA, and graphviz.org.
\\Email: s.c.n@ieee.org
\item
Carlos Scheidegger is with the University of Arizona, USA.
\\Email: cscheid@email.arizona.edu
\item
Simon Urbanek is with AT\&T Labs in Bedminster, USA.
\\Email: urbanek@research.att.com
\item
Gordon Woodhull is with AT\&T Labs in Bedminster, USA.
\\Email: gordon@research.att.com
}

@@

%% Abstract section.
\abstract{ Consider the emerging role of data science teams embedded in
  larger organizations. Individual analysts work on loosely related problems,
  and must share their findings with each other and the organization at large,
  moving results from exploratory data analyses (EDA)
  into automated visualizations, diagnostics and reports deployed for wider consumption.
  There are two problems with the current
  practice. First, there are gaps in this workflow: EDA is performed
  with one set of tools, and automated reports and deployments with another.
  Second, these environments often assume a single-developer perspective,
  while data scientist teams could get much benefit from easier sharing
  of scripts and data feeds, experiments, annotations, and automated
  recommendations, which are well beyond what traditional version control
  systems provide. We contribute and justify the following three
  requirements for systems built to support current data science teams
  and users: \emph{technology transfer}, \emph{coexistence}, and
  \emph{discoverability}. In addition, we contribute the design and
  implementation of RCloud, a system that supports the requirements of
  collaborative data analysis, visualization and web deployment.
  RCloud has been in active use for over two years by about 100 active users.
  We report on interviews with some of these users, and
  discuss the design decisions, tradeoffs and limitations.
  % comparing RCloud to other current proposals.
} % end of abstract

@@

%% Uncomment below to include a teaser figure.
\teaser{
  \centering
  \includegraphics[width=.80\linewidth]{fig/teaser/teaser.pdf}
  % \vspace{-.5em}
  \caption{
    An overview of features in RCloud. RCloud supports an environment in which a large number of loosely-related problems in data and visual analytics are solved by a small number of \emph{hackers} and \emph{scripters}. In such a shared environment, \emph{discoverability} is a concern. RCloud supports search, annotation, recommendation, and commenting for all notebooks, and provides an overview where users can \emph{browse} popular and recent analyses. When problems and data sources change frequently, deployment can become very costly; RCloud supports \emph{transparent and automatic} deployment of analyses as web pages and web services, allowing a seamless transition from data exploration work to production web services.
  }
% \vspace{-.5em}
}

### 2015-06-25 system.tex
fixups

Implicit signaling of interest is supported by keeping click-through
counts~\cite{Joachims:2005:AIC} and execution counts. In addition to
these standard techniques, we hope to apply static and dynamic code
analysis to infer fine-grained information about relationships,
for example, which packages and data sets often appear together,
in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.

>>>>

Implicit signaling of interest is supported by keeping click-through
counts~\cite{Joachims:2005:AIC} and execution counts. In addition to
these standard techniques, we hope to apply static and dynamic code
analysis to infer fine-grained information about relationships,
for example, which packages and data sets often appear together,
% in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.
in the style of Codex.~\cite{Fast:2014:ECS}.
